### 2016-10-25 doc/publications/plos16/plos16.tex
Fix some semantics

\newcommand{\cursedforest}{{\sc CursedForest}}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\cursedforest - A random forest implementation for ``big'' and ``wide'' data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1},
Piotr Szul\textsuperscript{2},
Stephanie Li\textsuperscript{3},
James Doecke\textsuperscript{3},
Nick Ellis\textsuperscript{4},
Robert Dunne\textsuperscript{5}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\textbf{2} Data61, CSIRO, Brisbane, QLD, Australia
\\
\textbf{3} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\textbf{4} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\\
\textbf{5} Data61, CSIRO, Sydney, NSW, Australia
\\
\bigskip

@@

\end{flushleft}
\tableofcontents
\clearpage
\section{Notes}
\begin{itemize}
\item   I think the structure of the first section should be
  \begin{itemize}
  \item  convergence results from Biau 2012  (done)
  \item consider  \cite{Segal.2004} examples
  \item simulations reproduced from Genuer et al 2010 (no -- we need classification examples)
  \item check Piotr's simulation  (done)
  \item run sparkML or CursedForest on the  simulation data 
  \item we extend the simulations to many more variables 
  \item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
    \ref{table:smaller_data} -- On it, Aidan
  \item variable  selection may succeed where accurate prediction does not
  \item discussion of the effect of the ntree and mtry parameters for variable selection with many noise variables
  \item   perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form (step, ramp) we can recover
    the significant variables in the presence of a very large number of noise variables
  \item in the case of genomic data, our functions of interest are more likely to be stop or ramp functions rather than harmonic
    functions. So we have some hope.
  \end{itemize}
\item check out 
  \begin{itemize}
  \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). (done)
  \item \cite{Goldstein.et.al.2011}
notes that the 0, 1, 2 encoding of alleles is an additive model i.e. each additional minor allele increases the effect 
\begin{verbatim}
Different genetic effects
Type	Mechanism	Partition
Additive	Each additional minor allele increases variation	0, 1, 2
Dominant	Presence of at least 1 minor allele increases variation	0, 1/2
Recessive	Two minor alleles needed for variation	0/1, 2
Heterosis	Heterozygote leads to variation	0/2, 1
\end{verbatim}

@@

% Please keep the abstract below 300 words
\section*{Abstract}
The abstract.

@@

\section{Introduction}
\marginpar{change parallelisation to parallelization}

@@

Identifying the relevant and disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs
(SNPs) by applying linear or logistic regression analysis to each SNP separately~\cite{CCC2007}.  It has since been
demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
is inadequate~\cite{Manolio2009}.  
\marginpar{is Complex Trait a thing? why capitals?}
Complex Trait analysis has demonstrated that in order to explain the overall
heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.

\marginpar{decison trees capture interaction -- there is no main effect term. RF inherit this property. I dont think it
  has anything to do with boostrapping, the point of which is to ensure that successive trees are not correlated.}
Random Forests~\cite{Breiman2001} (RF) are able to capture interactions between features while still allowing to
prioritise SNPs for their contribution to disease.  The joint interaction is achieved by appling the technique of
bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example,
by summing the number of times a variable is used to split a node, over all nodes and all trees.  RF are well suited for
processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. 
Secondly, RF are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
separate trees on different processors and combine the trees.

@@

However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed by
these methods.  The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) or
else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  In the case of
random forests this means that the computations for each node of a tree can be handed off to separate processors.

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forest specifically designed to cater for
``big'' (many samples) and ``wide'' (many features) data sets. \cursedforest\ is capable of parallelising the split for
each node in a tree thereby handling millions of features, such as in the 1000 Genomes Project data, which consists
of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also utilising Spark to read in and
manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small
datasets where multi-threading generally performs well. 
\marginpar{``performant'' -- a person who performs, an actor or celebrant }
 \marginpar{Piotr this is your 20 trees has the important
  feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be
  a result section.}
Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.

@@

A Spark application runs on a ``driver'' node, and this node
divides and distributes work out to the many ``worker'' nodes, or ``executors''.

@@

Here we extend this toolkit to encompass supervised learning using random forest.  Spark ML's RF implementation is not
able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest
dimensionality.  Although Spark ML can build a RF model on a subset of the data (chromosome 1~and~2), the time taken is
excessive due to the huge amount of data being aggregated and processed by the single driver node during intermediate
stages of building the model (see Result~\ref{comp}).  This unbalanced work load where the driver node becomes the
bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per
executor also increases with dimensionality due to the data types Spark ML uses, which we elaborate on in the
discussion.

@@

%Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.  
%Therefore, these historically successful solutions are not catering for the big data demands that especially due to WGS are becoming more prevalent. Here, classification or feature importance analysis needs to be performed on datasets with a much larger number of features~\cite{Yano2016}. 
%For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
%This dataset consumes close to a terabyte of disk space.  
%We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
%framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
%Using VariantSpark, we successfully built a k-means model on the full
%$2500 \times 80$ million matrix to cluster individuals by
%their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).

%Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
%Although Spark ML can build a RF model on
%a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
%and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
%unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
%scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
%Spark ML uses, which we elaborate on in the discussion.
%
%We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
%Unlike
%the standard Apache implementation, \cursedforest\ is able to easily process data with millions of dimensions without
%excessive resource requirements on the driver or worker nodes.
%DENIS Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.

\marginpar{Needs to be rewritten once we actually have the sections} 
In the first section we explore some of the
properties of random forests, demonstrating that any potential flaws in \cursedforest\ are an inherent property of Random
Forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of \cursedforest\ in
respect to the dimensionality of the data, building an RF model on whole-genome data from the 1000 Genomes Project. Here
we report on the accuracy (in terms of the ARI)
\marginpar{what is ARI?}
 and the error.  Finally, given the role different parameter values can
play in model construction, we explore the effect tuning these parameters can have on the prediction accuracy of the
model.

@@

Random Forests~\cite{Breiman2001} (RF) apply the technique of bootstrap aggregation or bagging to decision trees.  The training
data is independently drawn from the joint distribution of $(X,Y)$ and comprises $n$ $(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$.
$X$ and $\theta_b$ for $b=1,\ldots,B$ are i.i.d. random vectors and a tree is grown on each of these $\theta_b$  samples 
($X$ is randomly sampled $B$ times with replacement and a tree is grown on each of these $B$ samples). 
The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}=  \arg \max_q \left(\sum_b I(h(x;\theta_b)=q)\right).
\end{equation*}

@@

The benefits of random forests models are:
  \begin{itemize}
  \item they give an out-of-bag estimate of model accuracy, by using the samples not selected for the bootstrapped sample
    as an independent test set for each tree;
  \item they provide a measure of variable importance, for example by summing the number of times a variable is used to
    split a node, over all nodes and all trees. More complex measures of variable importance are also used;
  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
    variables. This would be difficult as the trees as split on random subsets of variables).
  \end{itemize}

Since the original implementation of random forests there have been a number of developments leading to a better understanding of
the algorithm. 
\begin{itemize}
\item random forests are resistant against overfitting. However it is not true that they will not overfit,
  ~\cite{Segal.2004} gives examples of overfitting in the case where:
  \begin{itemize}
  \item a single tree is the most appropriate model; and
  \item the variables are hightly correlated so that the boostrap resampling does not give rise to uncorrelated trees.
  \end{itemize}
However, keeping these difficult cases in mind, RF have a strong claim to being resistant to overfitting.

\item \cite{Strobl.et.al.2007} demonstrates that the RF algorithm is subject to a bias in variable selection via two mechanisms:
  \begin{enumerate}
  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
    bias. A uninformative variable with a large number of levels may be selected over a more informative variable with
    fewer levels;
  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$  of the data without replacement seems to be an
    effective stragey for removing the problem.
  \end{enumerate}
\end{itemize}

\cite{Strobl.et.al.2007} introduce changes to the RF algorithm to fix 1) and 2).  However the level effect will be
apparent in the CF algorithm as well.  However, there are many examples, particularly in genomics, where we have data
sets that are
\begin{itemize}
\item very wide;
\item all variables have the same number of levels (for example, measuments at different points along the genome);
\item the hypothesized functional relationship is of a form amenable to a decision tree.
\end{itemize}

@@

Subject to certain conditions, RF will not overfit even in the  $p > n$ case. 
Here we demonstrate that this may hold true in extreme cases where $p \gg n$.

\marginpar{why is this relevant, i.e. if this is special to VariantSpark it should be explained better.}
\subsubsection{variable selection}
\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
data.
\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of RF to handle 
GWAS data. What do we need to do in this area? What about linkage diseqelibrium?}

@@

\subsection{Parameter settings}
We consider the parameter setting for the RF algorithm. We use the R terms from the package RandomForests
\cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler. We incorporate the 
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience.

\begin{itemize}
\item \texttt{ntree} the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
  \marginpar{can we do proximity?}  \cite{Liaw.and.Weiner.2002} suggest that a high \texttt{ntree} is necessary to get stable
  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
  the ranking of the importances is quite stable.  In addition, we note that it is possible for a RF model to have a very poor
  fit, and be useless for prediction, and still have a useful ranking of variable importance.
\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \texttt{mtry} may give
  better performance.
\item  For classification problems where the class frequencies are extremely unbalanced it may be necessary to change the prediction
  rule \marginpar{can we do this}
\item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest. The SparkML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
  example\footnote{The documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html}.}.

  In the ATCG example the RandomForests code produced trees with depths of 30+. We note that Random forests are a way of averaging
  multiple deep decision trees, trained on different parts of the same training set, with the goal of reducing the
  variance. \url{https://en.wikipedia.org/wiki/Random_forest} and refs therein.  
\marginpar{are we including this  example?  Supplementary info?}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \texttt{ntree}, in
%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
%run, the ranking of the variables will stabilise with growing \texttt{ntree}.
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
complex trait of multiple interacting features.

@@

Figure~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
(mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
the weight originally assigned to them (dotted line). Further increasing $mtry$ improves the predicted importance
measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
15 remaining variables from the list.

Figure~\ref{figure:synth}b illustrates that increasing $mtry$ also improves the OOB (out of bag) classification error.
%Piotr can you please explain the remaining error on the synthetic data
By $mtry=0.8$ the error is converging to X, which is due to Y.

@@

\begin{figure}[tbhp]
\begin{tabular}{ll}
a)& b)\\
\includegraphics[totalheight=6cm]{./figs/var_oder.png} & 
\includegraphics[totalheight=6cm]{./figs/var_oob.png} \\
\end{tabular}
\captionof{figure}{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for
    random forest with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
  importance of top 20 variabels normalized to the sum of the linear model coefficients. {\bf Others} represents
  aggregated importance of variables not present in the linear model.  {\bf b)} OOB (out of bag) classification error
  estimate as funtion of mtry fraction.}
\label{figure:synth}
\end{figure}

@@

In Table~\ref{table3}, we display the top 10 features, ranked by their Gini index in the Random Forest model.

>>>>

\newcommand{\cursedforest}{{\sc CursedForest}}
\newcommand{\mtry}{{\texttt mtry}}
\newcommand{\ntree}{{\texttt ntree}}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1},
Piotr Szul\textsuperscript{2},
Stephanie Li\textsuperscript{3},
James Doecke\textsuperscript{3},
Nick Ellis\textsuperscript{4},
Robert Dunne\textsuperscript{5}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\textbf{2} Data61, CSIRO, Brisbane, QLD, Australia
\\
\textbf{3} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\textbf{4} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\\
\textbf{5} Data61, CSIRO, Sydney, NSW, Australia
\\
\bigskip

@@

\end{flushleft}
\tableofcontents
\clearpage
\section{Notes}
\begin{itemize}
\item I think the structure of the first section should be
  \begin{itemize}
  \item convergence results from Biau 2012  (done)
  \item consider  \cite{Segal.2004} examples
  \item simulations reproduced from Genuer et al 2010 (no -- we need classification examples)
  \item check Piotr's simulation  (done)
  \item run sparkML or CursedForest on the  simulation data 
  \item we extend the simulations to many more variables 
  \item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
    \ref{table:smaller_data} -- On it, Aidan
  \item variable  selection may succeed where accurate prediction does not
  \item discussion of the effect of the \ntree\ and \mtry\ parameters for variable selection with many noise variables
  \item   perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form (step, ramp) we can recover
    the significant variables in the presence of a very large number of noise variables
  \item in the case of genomic data, our functions of interest are more likely to be stop or ramp functions rather than harmonic
    functions. So we have some hope.
  \end{itemize}
\item check out 
  \begin{itemize}
  \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). (done)
  \item \cite{Goldstein.et.al.2011}
notes that the 0, 1, 2 encoding of alleles is an additive model i.e. each additional minor allele increases the effect 
\begin{verbatim}
Different genetic effects
Type	Mechanism	Partition
Additive	Each additional minor allele increases variation	0, 1, 2
Dominant	Presence of at least 1 minor allele increases variation	0, 1/2
Recessive	Two minor alleles needed for variation	0/1, 2
Heterosis	Heterozygote leads to variation	0/2, 1
\end{verbatim}

@@

% Please keep the abstract below 300 words
\section{Abstract}
The abstract.

@@

\section{Introduction}

@@

Identifying the relevant and disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs
(SNPs) by applying linear or logistic regression analysis to each SNP separately~\cite{CCC2007}.  It has since been
demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
is inadequate~\cite{Manolio2009}.  
Complex trait analysis has demonstrated that in order to explain the overall
heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.

\marginpar{decison trees capture interaction -- there is no main effect term. Random forests inherit this property. I dont think it
  has anything to do with boostrapping, the point of which is to ensure that successive trees are not correlated.}
Random forests~\cite{Breiman2001} are able to capture interactions between features while still allowing to
prioritise SNPs for their contribution to disease.  The joint interaction is achieved by appling the technique of
bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example,
by summing the number of times a variable is used to split a node, over all nodes and all trees.  Random forests are well suited for
processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. 
Secondly, random forests are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
separate trees on different processors and combine the trees.

@@

However, the use of traditional compute infrastructure limits the parallelization strategies that can be employed by
these methods.  The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) or
else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelization strategies.  In the case of
random forests this means that the computations for each node of a tree can be handed off to separate processors.

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater for
``big'' (many samples) and ``wide'' (many features) data sets. \cursedforest\ is capable of parallelising the split for
each node in a tree thereby handling millions of features, such as in the 1000 Genomes Project data, which consists
of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also utilising Spark to read in and
manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools even on small
datasets where multi-threading generally performs well. 
\marginpar{``performant'' -- a person who performs, an actor or celebrant }
 \marginpar{Piotr this is your 20 trees has the important
  feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be
  a result section.}
Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.

@@

A Spark application runs on a ``driver'' node and distributes tasks to the many ``worker'' nodes, or ``executors''.

@@

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest implementation is not
able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest
dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome 1~and~2), the time taken is
excessive due to the huge amount of data being aggregated and processed by the single driver node during intermediate
stages of building the model (see Result~\ref{comp}).  This unbalanced work load where the driver node becomes the
bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per
executor also increases with dimensionality due to the data types Spark ML uses, which we elaborate on in the
discussion.

@@

%Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
%Therefore, these historically successful solutions are not catering for the big data demands that especially due to WGS are becoming more prevalent. Here, classification or feature importance analysis needs to be performed on datasets with a much larger number of features~\cite{Yano2016}. 
%For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
%This dataset consumes close to a terabyte of disk space.  
%We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
%framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
%Using VariantSpark, we successfully built a k-means model on the full
%$2500 \times 80$ million matrix to cluster individuals by
%their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering and -1 random clustering).

%Spark ML's random forest implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
%Although Spark ML can build a random forest model on
%a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
%and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
%unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
%scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
%Spark ML uses, which we elaborate on in the discussion.
%
%We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
%Unlike
%the standard Apache implementation, \cursedforest\ is able to easily process data with millions of dimensions without
%excessive resource requirements on the driver or worker nodes.
%DENIS Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.

\marginpar{Needs to be rewritten once we actually have the sections} 
In the first section we explore some of the
properties of random forests, demonstrating that any potential flaws in \cursedforest\ are an inherent property of random
forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of \cursedforest\ in
respect to the dimensionality of the data, building a random forest model on whole-genome data from the 1000 Genomes Project. Here
we report on the accuracy (in terms of the ARI)
\marginpar{what is ARI?}
 and the error.  Finally, given the role different parameter values can
play in model construction, we explore the effect tuning these parameters can have on the prediction accuracy of the
model.

@@

Random forests~\cite{Breiman2001} (RFs) apply the technique of bootstrap aggregation or bagging to decision trees.  The training
data is independently drawn from the joint distribution of $(X,Y)$ and comprises $n$ $(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$.
$X$ and $\theta_b$ for $b=1,\ldots,B$ are i.i.d. random vectors and a tree is grown on each of these $\theta_b$ samples 
($X$ is randomly sampled $B$ times with replacement and a tree is grown on each of these $B$ samples). 
The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;\theta_b)=q)\right).
\end{equation*}

@@

The benefits of random forest models are:
  \begin{itemize}
  \item they give an out-of-bag estimate of model accuracy, by using the samples not selected for the bootstrapped sample
    as an independent test set for each tree;
  \item they provide a measure of variable importance, for example by summing the number of times a variable is used to
    split a node, over all nodes and all trees. More complex measures of variable importance are also used;
  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
    variables. This would be difficult as the trees as split on random subsets of variables).
  \end{itemize}

Since the original implementation of random forests, there have been a number of developments leading to a better understanding of
the algorithm. 
\begin{itemize}
\item random forests are resistant to overfitting. However it is not true that they will not overfit,
  ~\cite{Segal.2004} gives examples of overfitting in the case where:
  \begin{itemize}
  \item a single tree is the most appropriate model; and
  \item the variables are hightly correlated so that the boostrap resampling does not give rise to uncorrelated trees.
  \end{itemize}
However, keeping these difficult cases in mind, random forests have a strong claim to being resistant to overfitting.

\item \cite{Strobl.et.al.2007} demonstrates that the random forest algorithm is subject to a bias in variable selection via two mechanisms:
  \begin{enumerate}
  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
    bias. A uninformative variable with a large number of levels may be selected over a more informative variable with
    fewer levels;
  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$ of the data without replacement seems to be an
    effective stragey for removing the problem.
  \end{enumerate}
\end{itemize}

\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  However the level effect will be
apparent in the CF algorithm as well.  However, there are many examples, particularly in genomics, where we have data
sets that are
\begin{itemize}
\item very wide;
\item all variables have the same number of levels (for example, measuments at different points along the genome);
\item the hypothesized functional relationship is of a form amenable to a decision tree.
\end{itemize}

@@

Subject to certain conditions, random forests will not overfit even in the $p > n$ case. 
Here we demonstrate that this may hold true in extreme cases where $p \gg n$.

\marginpar{why is this relevant, i.e. if this is special to VariantSpark it should be explained better.}
\subsubsection{variable selection}
\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
data.
\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of random forests to handle 
GWAS data. What do we need to do in this area? What about linkage diseqelibrium?}

@@

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R terms from the {\sc randomForest} package
\cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler. We incorporate the 
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience.

\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
  \marginpar{can we do proximity?} \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable
  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
  the ranking of the importances is quite stable.  In addition, we note that it is possible for a random forest model to have a very poor
  fit, and be useless for prediction, and still have a useful ranking of variable importance.
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance.
\item For classification problems where the class frequencies are extremely unbalanced it may be necessary to change the prediction
  rule \marginpar{can we do this}
\item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest. The Spark ML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
  example\footnote{The documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html}.}.

  In the ATCG example the RandomForests code produced trees with depths of 30+. We note that random forests are a way of averaging
  multiple deep decision trees, trained on different parts of the same training set, with the goal of reducing the
  variance. \url{https://en.wikipedia.org/wiki/Random_forest} and refs therein.  
\marginpar{are we including this example?  Supplementary info?}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in
%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
%run, the ranking of the variables will stabilise with growing \ntree.
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
complex trait of multiple interacting features.

@@

Figure~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
(\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
15 remaining variables from the list.

Figure~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB (out of bag) classification error.
%Piotr can you please explain the remaining error on the synthetic data
By $mtry=0.8$ the error is converging to X, which is due to Y.

@@

\begin{figure}[tbhp]
\begin{tabular}{ll}
a)& b)\\
\includegraphics[totalheight=6cm]{./figs/var_oder.png} & 
\includegraphics[totalheight=6cm]{./figs/var_oob.png} \\
\end{tabular}
\captionof{figure}{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for
  random forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
  importance of top 20 variabels normalized to the sum of the linear model coefficients. {\bf Others} represents
  aggregated importance of variables not present in the linear model.  {\bf b)} OOB (out of bag) classification error
  estimate as funtion of \mtry\ fraction.}
\label{figure:synth}
\end{figure}

@@

In Table~\ref{table3}, we display the top 10 features, ranked by their Gini index in the random forest model.
